Value-Suppressing Uncertainty Maps

### 2017-09-12 paper/main.tex
Check in for pilot fixes

\section{Introduction}
certainty is an inescapable component of collecting, presenting, and using data. A common goal in the communication of uncertainty is promoting \emph{uncertainty-aware decisions}: the audience should be aware of the risks and rewards of certain decisions, modulate their confidence in their conclusions, and perhaps refrain from making a decision at all if there is too much uncertainty.  A way that designers can contribute to this goal is by ensuring that uncertainty information is \emph{well-integrated} with the rest of the data. That is, it should be difficult to discount or ignore the uncertainty in a dataset.

@@

In this paper, we contribute \textbf{Value-Suppressing Uncertainty Maps} (VSUMs) for integrating data and uncertainty information in visualizations. VSUMs intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. Compared to a traditional two-dimensional bivariate map, with differing outputs for each combination of value and uncertainty, VSUMs can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. \figref{fig:example} shows examples of both a more traditional bivariate map and a VSUM.

We describe the motivations behind the use of VSUMs, examples of their utility for decision-making under uncertainty, and assess VSUMs in a crowdsourced experiment. Our experimental results indicate that VSUMs create close integration between uncertainty and data, promoting rapid but cautious decision-making.

@@

Reviewing the state of the art in uncertainty visualization in the field, Greithe \ea~\cite{griethe2006visualization} and Brodlie \ea~\cite{brodlie2012review} present lists of potential techniques for conveying uncertainty in concert with data. Some of these options (interaction, animation, and sonification) are not applicable for static charts; even for dynamic charts, many users do not interact with charts in sufficient detail to recover uncertainty information~\cite{nyt2016}. While we acknowledge the potential utility of these techniques in uncertainty visualization (for instance, the animation used to convey sampling error in Hullman \ea's Hypothetical Outcome Plots~\cite{hullman2015hypothetical}), we limit the scope of our discussion to techniques which center on static charts.

There are a number of techniques for communicating values and uncertainty in one-dimensional statistical graphics~\cite{olston2002visualizing, wickham201140, neyman1937outline}. Techniques such as error bars, confidence bands, and violin plots~\cite{hintze1998violin} use a separate position channel to encode uncertainty. The techniques in this paper also apply to 2D visualizations such as heatmaps. To our knowledge, there are no established standards of visualizing uncertainty in heatmaps.

\strategyFig

To support our goal of making uncertainty information well-integrated with data, we focus on the techniques of \emph{juxtaposition} (where uncertainty is presented alongside the data), \emph{superposition} (where additional glyphs are overlaid on top of the data), and \emph{explicit encoding} (where an additional visual channel is used to encode the uncertainty). Gleicher \ea~\cite{gleicher2011visual} explore these three strategies from the perspective of affording \emph{comparisons}; here, we examine them from the perspective of incorporating measures of uncertainty. \figref{fig:strategies} shows simple examples of each of these three techniques for presenting both uncertainty and value.

\subsection{Strategies for Encoding Uncertainty}

\emph{Juxtaposition} creates two explicit visualizations: an ``uncertainty map'' and a ``data map.'' MacEachren~\cite{maceachren1992visualizing} portrays uncertainty visualization as the design problem of how to \emph{unify} these two maps, and so this explicit separation can be counterproductive to the goal of information fusion. Juxtaposition also creates an additional search task for viewers, who must register hotspots in one map with corresponding features of the other map. Rather than complete this task, viewers may choose to ignore the uncertainty information altogether. This ignoreability is a key deficiency for information fusion tasks. Nevertheless, especially in cases where the quantification of uncertainty is a separate analysis process from the measuring of data values, juxtaposition is a common design tactic for conveying uncertainty~\cite{moritz2017trust}.

\emph{Superposition}, or the overlay of glyphs representing uncertainty on top of a data map, is another common strategy. Error bars are a common example of this strategy: a mark indicating measurement error is superimposed atop a mark representing a point value. Ware's~\cite{ware2009quantitative} \emph{textons} are an example of this strategy applied to thematic maps: monochrome glyphs that can encode an ordinal variable, or a binned quantitative variable. Greithe \ea~\cite{griethe2006visualization} and MacEachren~\cite{maceachren1992visualizing,maceachren1998visualizing} present other examples of superimposed uncertainty in thematic maps, including overlaid grids with ``fuzzy'' lines to indicate uncertainty, or overlaid treemaps with coarser and coarser resolution in uncertain regions. Costs to this approach include obscuring the underlying data map (which can cause uncertainty information to dominate the display~\cite{brodlie2012review}), potential visual interactions that can impair the legibility of glyphs (for instance, simultaneous contrast effects), and the necessity of performing spatial binning to ensure that there are a discrete number of uncertainty glyphs. How this binning is performed can introduce undesirable artifacts on the resulting visualization, suppressing important signals or highlighting spurious ones~\cite{battersby2016shapes}.

\emph{Explicit encoding} directly represents uncertainty with a visual channel, simultaneously with data. VSUMs employ this strategy. In addition to overcoming the drawbacks of the other techniques mentioned in this section, we believe that explicit encoding represents the most general form of the solution of how to fuse uncertainty information with data. The most pressing design decisions for explicit encoding are 1) which visual variables to use to encode uncertainty and 2) how to design bivariate maps which afford the encoding of both value and uncertainty. We deal with these two questions in greater detail below.

@@

Foundational work in psychophysics and graphical perception make use of the concept of a ``just noticeable difference'' (JND)\cite{cleveland1984graphical}. A 50\% JND is a difference between two stimuli that is detectable 50\% of the time. For scales and legends, we would want a significantly higher standard of unambiguity. That is, the different levels of visual variables used in visualizations should be readily and reliably distinguishable. Complicating this goal is that a JND as measured in a controlled environment with cooperative participants can be considerably lower than JNDs measured in the sorts of heterogeneous conditions found in real environments~\cite{szafir2014adapting}.

@@

@@

Dunn~\cite{dunn1989dynamic} attempts to circumvent these issues by dynamically allocating bins to bivariate maps. As the number of distinct, separable colors is limited, the dynamic approach allocates bins to regions of the highest utility. For instance, placing the dividing line between bins in areas aligned with a trend line allows the color to encode information about the sign of the residual with respect to each variable. This insight, that the bins in a bivariate map are a limited resource and so should be ``spent'' wisely, drives the creation of VSUMs, which allocate more bins in a bivariate map to regions with less uncertainty.

\section{Value-Suppressing Uncertainty Maps}

Value-Suppressing Uncertainty Maps (VSUMs) are a technique for creating bivariate maps of data \emph{value} and \emph{uncertainty}. These maps circumvent the issues in the creation of traditional bivariate maps by making the following assumptions:

\begin{enumerate}
	\item \textbf{Designers have a limited budget of discriminable marks.} That is, viewers cannot easily or reliably discriminate between an infinite number of categories. Rather than accept this perceptual ambiguity, designers ought to \emph{alias} together marks into a finite, but sufficiently discriminable, set.
	\item \textbf{Uncertain values should have less weight than certain values.} Weight, in this case, might mean literal visual weight, or impact on the decision-making process.
\end{enumerate}

%MC this paragraph is a little iffy, and sort of "the lady doth protest too much" territory. we might be able to get away with "people bin colors all the time. and in fact, you really should be binning your colors anyway."
The first assumption seems to contradict Mackinlay's \emph{expressiveness principle}~\cite{mackinlay1986automating}, where the visual encoding of the data should reflect the composition of the data. For instance, if my data are a collection of measurements of some continuous variable, then it would make sense to encode these data using a continuous \emph{visual} variable, which would then mean an arbitrary number of potential marks. For instance, a bar can have an arbitrary, non-integer height. However, for many visualizations, the perceptual ambiguity caused by having marks that cannot be reliably disambiguated dominates the gains from having a visual encoding that maps readily to the backing data. Padilla \ea~\cite{padilla2017evaluating} present one such example, where binned heatmaps result in higher task accuracy than continuous heatmaps.

@@

If both of these assumptions hold, then it follows that the designer of a bivariate map should allocate more mark types to certain values, and fewer mark types to uncertain values. VSUMs codify this decision by \emph{reducing the number of mark categories for representing value as uncertainty increases.} This means that data encoded using a VSUM will make visible only the largest of differences in uncertain data, but highlight comparatively small differences in value when uncertainty is low. VSUMs therefore act as both a filtering mechanism (in that values with too much uncertainty are all mapped to the same glyph), as well as an implicit test of effect size (in that smaller and smaller changes in value are visible as uncertainty decreases). This strategy of dampening the low quality or high uncertainty values in maps in order to focus on more informative regions has measurable benefits, including the removal of statistically spurious visual patterns, and the highlighting of regions of interest~\cite{correll2017surprise}.

@@

The general VSUM construction algorithm is as follows:

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In this paper, this variable is position along some established color scale.
	\item \textbf{Select a visual variable to encode uncertainty}. We prefer variables that have a somewhat high semiotic connection with uncertainty (such as lightness and size), but still have a reasonable number of perceptually discriminable layers (as opposed to other, more intuitive channels such as blur or sketchiness). In most cases in this paper, we encode uncertainty as interpolation to white or as glyph size.
	\item \textbf{Select a discriminability threshold}. Choose some minimum distance, in a perceptual space, between marks. For many visual variables (such as length or hue), the perceptual discriminability between marks has been assessed in prior literature, and so the only step is to choose a value in terms of the just noticeable distance (JND). For less common visual variables, a perceptual model may be built empirically (as in Demiralp \ea~\cite{demiralp2014learning}).
	\item \textbf{Choose how quickly categories \emph{degrade}}. That is, if there are $n$ bins at some uncertainty level $m$, how many bins should there be when uncertainty increases to level $m+1$? In \figref{fig:example}, this function is $\left \lfloor n / 2 \right \rfloor$, meaning that the number of value bins decreases by half for each level of uncertainty bins. If there is too much degradation, then there are fewer levels of uncertainty. If there is no degradation, the resulting map is just a traditional bivariate map.
	\item \textbf{Choose the largest number of initial categories $n$ such that the discriminability threshold is met.} For each pair of marks in the map, perform a pairwise comparison of discriminability. If no marks are perceptually ambiguous by the standards of your threshold, the map is complete. Otherwise, decrement $n$.
\end{enumerate}

\figref{fig:flow} shows this process in detail, and how it can be used to generate a bivariate map for presenting uncertainty. JavaScript code for generating VSUMs given arbitrary D3~\cite{bostock2011d3} quantitative scale functions is available at \url{URL-REMOVED-FOR-REVIEW}.

Correll \ea~\cite{correll2015layercake,correll2011visualizing} use a precursor of VSUMs in their LayerCake genomics visualization tool, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``confidence fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both cases, the property of aliasing is \emph{ad hoc}, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUMs makes the choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

@@

There are multiple choices that designers must make before creating a VSUM. In particular, they must choose 1) which visual channels map to value and uncertainty, 2) a threshold of perceptual distinguishability (and, implicitly, a corresponding metric), and 3) a degradation function to determine how quickly uncertain values are aliased.

@@

How quickly the number of bins decreases is another parameter of VSUMs. If this degradation is too fast, then we may only have a few discriminable levels of uncertainty. If this degradation is too slow, then we can be overwhelmed with colors. For instance, if we only decrement by one bin per level of uncertainty, there would be $1,485$ different colors in a VSUM with a tolerance threshold of $1$ unit in CIELAB space, far too much complexity for a discrete color mapping. In the examples in this paper, we halve the number of bins at each increasing level of uncertainty. When the number of initial bins at the highly certain region is a power of $2$ (as in Fig. \ref{fig:example}), this imparts a tree structure to the resulting VSUM. As uncertainty increases, bins in the previous level map to one and only one ``parent'' in the next uncertainty level, resulting in fewer crossings of color boundaries among rows.

@@

We present a set of examples showing how VSUMs can be applied to realworld datasets. The decision of VSUMs to allocate color bins asymmetrically amongst value and uncertainty regions affords greater discriminability in regions of the chart that warrant closer scrutiny, while discouraging exploration of regions with noisy or unreliable signals.

@@

Using a public data set from the U.S. Bureau of Transportation Statistics~\cite{bts}, we created a heatmap showing average flight delay for different times of the day and days of the week for U.S. carriers in January 2017. We use standard error $\left(\sigma / \sqrt{n}\right)$ as our measure of uncertainty. Starting from an minimum discriminability threshold of $18$ units in CIELAB space, we created two alternate heatmaps of this delay information. \figref{fig:airline2d} shows a traditional bivariate map, whereas \figref{fig:airlineVsum} shows a VSUM generated under the same constraints.

Both maps illustrate a similar temporal trend: flight delays are shorter towards the beginning of the day, increasing on average over time. The traditional bivariate map affords only a coarse examination of this trend: two visible ``blocks'' of color, early in the day, and later in the afternoon. The VSUM, in contrast, has sufficient color resolution to show a quasi-linear trend. The additional color resolution for uncertainty information in VSUMs also promotes caution when considering outliers: the small number of flights on Saturdays, or departing late at night, make standard error for these delays relatively high, and the connected values more suspect than the rest of the chart.

@@

Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV). Next generation sequencing (NGS) allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) affords the analysis of only a few, large hot spots. Extending the ``confidence fog'' metaphor used in viral genome browsers like LayerCake~\cite{correll2015layercake}, VSUMs (\figref{fig:viralVsum}) enable biologists to explore interesting patterns while not being distracted by noise. Smaller scale ``hot spots'' that are mapped to the same magenta color in the traditional map are visible as bright orange lines. Pink hot spots towards the end of the genome (around reference nucleotide $9,500$) that could be plausibly considered potential regions of interest in the 2D map, are encoded as highly uncertain and of ambiguous value in the VSUMs. In fact, these regions have systematically low coverage; the lack of reliable data for these regions should discourage analysts from giving their apparent pattern equal consideration from the  stronger signals earlier in the genome.

@@

We performed a crowdsourced experiment on Amazon's Mechanical Turk to evaluate the effectiveness of VSUMs for integrating uncertainty and value information in visualizations. This focus on integration meant that we limited our experimental tasks to scenarios where the participants needed to consider both value and uncertainty before making a decision. We gave participants two main tasks:

@@

For each task, participants saw one of three types of heatmaps. Either two \textbf{juxtaposed} maps side-by-side, with separate encodings for value and uncertainty, a traditional \textbf{2D bivariate} map, where value and uncertainty are orthogonally encoded, or a \textbf{VSUM}. \figref{fig:conditions} shows examples of these factor levels. Other factors included the size of the map (either a 4x4 or 8x8 grid), and (for the prediction task), the question framing (attack or defense). Secondary factors were evenly blocked across stimuli, resulting in a 3x2 within subjects design for the identification task, and a 3x2x2 within subjects design for the prediction task. Participants reported fatigue effects in piloting, and so we limited the number of replicates in this study: participants saw 18 stimuli (6 for each map type) for the identification task, and 12 (4 for each map type) for the prediction task.

In all cases, we generated color maps based on the requirement that no two colors in the resulting color scheme were closer than 18 units of Euclidean distance in CIELAB space. This threshold results in a $9$-color bivariate map and $15$-color VSUM for all color scales used in this study. Given the size of the marks used in our study, this threshold distance is sufficient for a predicted discriminability rate of over 95\%~\cite{stone2014engineering}. Other choices of threshold, and differing univariate scales, would result in different maps (see Fig.~\ref{fig:performance}).

@@

For the identification task, we gave participants bivariate representations of uncertainty, and asked them to identify a location meeting a given criteria, e.g., ``Click on the location with the greatest uncertainty.'' There were three potential criteria, mapping to the three vertices of a VSUM: 1) high value, low uncertainty, 2) low value, low uncertainty, and 3) high uncertainty. Each map contained four locations, one of each combination of maximum and minimum value and uncertainty. The correct answer was always one of these locations. Other locations in the grid were created pseudo-randomly, with less extreme ranges of uncertainty and probability. We measured performance both in terms of accuracy (did the participant select the correct point) as well as response time.

@@

%Show:
% 1) Juxtaposition makes people ignore/underweight uncertainty (even when they don't screw up the search task)
% 2) VSUMs make people, surprise surprise, bad at distinguishing highly uncertain values
% 3) But, they make uncertainty better-integrated
%    i) lower weights on uncertain values in decision-making
%    ii) a priori better discrimination

@@

Across both the identification and prediction tasks, we had several hypotheses, stemming from our belief that VSUMs promote better \emph{integration} between uncertainty and value information, and also encourage \emph{caution} by highlighting the ambiguity or untrustworthiness introduced by uncertain data. In particular:
\begin{enumerate}
	\item Participants would be \textbf{faster and more accurate} when completing the identification task using a VSUM as compared to the other map types. Juxtaposed maps, by adding an additional search task to the lookup task, would have especially poor performance compared to VSUMs.
	\item Participants would choose targets with \textbf{less uncertainty} in the prediction task when using a VSUM compared to the other maps types. This would result in a tradeoff where they would also choose targets with \textbf{less expected value} than the other map types. Again, juxtaposed maps, by increasing the difficulty of retrieving both variables simultaneously, would highlight this difference.
\end{enumerate}

@@

For the identification task, our primary measures of performance were accuracy and response time. Our hypothesis was that VSUMs would exhibit better performance than the other charts.

We performed a one-way ANOVA investigating the effect of chart type (juxtaposed, full bivariate, or VSUM) on accuracy. We did not find a significant effect ($F(2,859) = 2.4$, $p=0.09$), although VSUMs had the highest accuracy. \figref{fig:accuracy1} illustrates this result.

@@

For the prediction task, where there is no dominating strategy, our primary measure was the distribution of guesses, in terms of both their average value, and their average uncertainty. A risky player would choose guesses with high value, regardless of the uncertainty of those points. A more conservative guesser might eschew high-risk, high-reward locations, resulting in a lower average value of guesses, but also lower uncertainty. Our hypothesis is that VSUMs, by communicating uncertainty information at the expense of value information, would promote more conservatives guesses than the other chart types.

>>>>

\section{Introduction}
Uncertainty is an inescapable component of collecting, presenting, and using data. A common goal in the communication of uncertainty is promoting \emph{uncertainty-aware decisions}: the audience should be aware of the risks and rewards of certain decisions, modulate their confidence in their conclusions, and perhaps refrain from making a decision at all if there is too much uncertainty.  A way that designers can contribute to this goal is by ensuring that uncertainty information is \emph{well-integrated} with the rest of the data. That is, it should be difficult to discount or ignore the uncertainty in a dataset.

@@

In this paper, we contribute \textbf{Value-Suppressing Uncertainty Palettes} (VSUPs) for integrating data and uncertainty information in visualizations.
VSUP intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. If traditional bivariate maps are a two-dimensional square, with differing outputs for each combination of value and uncertainty, VSUP can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. \figref{fig:example} shows examples of both a more traditional bivariate map and a VSUP.

We describe the motivations behind the use of VSUP, examples of their utility for decision-making under uncertainty, and assess VSUP in a crowdsourced experiment. Our experimental results indicate that VSUP create close integration between uncertainty and data, promoting rapid but cautious decision-making.

@@

To our knowledge, there are no established standards for visualizing uncertainty in heatmaps. Reviewing the state of the art in uncertainty visualization in the field, Greithe \ea~\cite{griethe2006visualization} and Brodlie \ea~\cite{brodlie2012review} present lists of potential techniques for conveying uncertainty in concert with data. Some of these options (interaction, animation, and sonification) are not applicable for static charts; even for dynamic charts, many users do not interact with charts in sufficient detail to recover uncertainty information~\cite{nyt2016}. While we acknowledge the potential utility of these techniques in uncertainty visualization (for instance, the animation used to convey sampling error in Hullman \ea's Hypothetical Outcome Plots~\cite{hullman2015hypothetical}), we limit the scope of our discussion to techniques which center on static charts.

%There are a number of techniques for communicating values and uncertainty in one-dimensional statistical graphics~\cite{olston2002visualizing, wickham201140, neyman1937outline}. Techniques such as error bars, confidence bands, and violin plots~\cite{hintze1998violin} use a separate position channel to encode uncertainty. The techniques in this paper also apply to 2D visualizations such as heatmaps.

%\strategyFig

%To support our goal of making uncertainty information well-integrated with data, we focus on the techniques of \emph{juxtaposition} (where uncertainty is presented alongside the data), \emph{superposition} (where additional glyphs are overlaid on top of the data), and \emph{explicit encoding} (where an additional visual channel is used to encode the uncertainty). Gleicher \ea~\cite{gleicher2011visual} explore these three strategies from the perspective of affording \emph{comparisons}; here, we examine them from the perspective of incorporating measures of uncertainty. \figref{fig:strategies} shows simple examples of each of these three techniques for presenting both uncertainty and value.

%\subsection{Strategies for Encoding Uncertainty}

%\emph{Juxtaposition} creates two explicit visualizations: an ``uncertainty map'' and a ``data map.'' MacEachren~\cite{maceachren1992visualizing} portrays uncertainty visualization as the design problem of how to \emph{unify} these two maps, and so this explicit separation can be counterproductive to the goal of information fusion. Juxtaposition also creates an additional search task for viewers, who must register hotspots in one map with corresponding features of the other map. Rather than complete this task, viewers may choose to ignore the uncertainty information altogether. This ignoreability is a key deficiency for information fusion tasks. Nevertheless, especially in cases where the quantification of uncertainty is a separate analysis process from the measuring of data values, juxtaposition is a common design tactic for conveying uncertainty~\cite{moritz2017trust}.

%\emph{Superposition}, or the overlay of glyphs representing uncertainty on top of a data map, is another common strategy. Error bars are a common example of this strategy: a mark indicating measurement error is superimposed atop a mark representing a point value. Ware's~\cite{ware2009quantitative} \emph{textons} are an example of this strategy applied to thematic maps: monochrome glyphs that can encode an ordinal variable, or a binned quantitative variable. Greithe \ea~\cite{griethe2006visualization} and MacEachren~\cite{maceachren1992visualizing,maceachren1998visualizing} present other examples of superimposed uncertainty in thematic maps, including overlaid grids with ``fuzzy'' lines to indicate uncertainty, or overlaid treemaps with coarser and coarser resolution in uncertain regions. Costs to this approach include obscuring the underlying data map (which can cause uncertainty information to dominate the display~\cite{brodlie2012review}), potential visual interactions that can impair the legibility of glyphs (for instance, simultaneous contrast effects), and the necessity of performing spatial binning to ensure that there are a discrete number of uncertainty glyphs. How this binning is performed can introduce undesirable artifacts on the resulting visualization, suppressing important signals or highlighting spurious ones~\cite{battersby2016shapes}.

%\emph{Explicit encoding} directly represents uncertainty with a visual channel, simultaneously with data. VSUP employ this strategy. In addition to overcoming the drawbacks of the other techniques mentioned in this section, we believe that explicit encoding represents the most general form of the solution of how to fuse uncertainty information with data. The most pressing design decisions for explicit encoding are 1) which visual variables to use to encode uncertainty and 2) how to design bivariate maps which afford the encoding of both value and uncertainty. We deal with these two questions in greater detail below.

@@

%TODO stuff that needs to get cited here:

% uncertainty vis surveys: kinkeldey2017evaluating,maceachren1992visualizing,brodlie2012review
% 2d uvis: griethe2006visualization, maceachren1992visualizing,maceachren1998visualizing
% 1d uvis: olston2002visualizing, wickham201140, neyman1937outline
% ignoreability: moritz2017trust

%Foundational work in psychophysics and graphical perception make use of the concept of a ``just noticeable difference'' (JND)\cite{cleveland1984graphical}. A 50\% JND is a difference between two stimuli that is detectable 50\% of the time. For scales and legends, we would want a significantly higher standard of unambiguity. That is, the different levels of visual variables used in visualizations should be readily and reliably distinguishable. Complicating this goal is that a JND as measured in a controlled environment with cooperative participants can be considerably lower than JNDs measured in the sorts of heterogeneous conditions found in real environments~\cite{szafir2014adapting}.

@@

%TODO cites, fill out section

% examples: roth2010value,ware2009quantitative

\subsection{Quantizing Color Scales}

%TODO cites, fill out section
% binning stuff: [data] battersby2016shapes, [legends] padilla2017evaluating,muller1979perception,dobson1973choropleth,dunn1989dynamic

@@

%Dunn~\cite{dunn1989dynamic} attempts to circumvent these issues by dynamically allocating bins to bivariate maps. As the number of distinct, separable colors is limited, the dynamic approach allocates bins to regions of the highest utility. For instance, placing the dividing line between bins in areas aligned with a trend line allows the color to encode information about the sign of the residual with respect to each variable. This insight, that the bins in a bivariate map are a limited resource and so should be ``spent'' wisely, drives the creation of VSUPs, which allocate more bins in a bivariate map to regions with less uncertainty.

\section{Value-Suppressing Uncertainty Palettes}

Value-Suppressing Uncertainty Palettes (VSUPs) are a technique for creating bivariate maps of data \emph{value} and \emph{uncertainty}.

@@

If both of these assumptions hold, then it follows that the designer of a bivariate map should allocate more mark types to certain values, and fewer mark types to uncertain values. VSUPs codify this decision by \emph{reducing the number of mark categories for representing value as uncertainty increases.} This means that data encoded using a VSUP will make visible only the largest of differences in uncertain data, but highlight comparatively small differences in value when uncertainty is low. VSUPs therefore act as both a filtering mechanism (in that values with too much uncertainty are all mapped to the same glyph), as well as an implicit test of effect size (in that smaller and smaller changes in value are visible as uncertainty decreases). This strategy of dampening the low quality or high uncertainty values in maps in order to focus on more informative regions has measurable benefits, including the removal of statistically spurious visual patterns, and the highlighting of regions of interest~\cite{correll2017surprise}.

@@

The general VSUP construction algorithm is as follows:

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In this paper, this variable is position along some established color scale.
\end{enumerate}

\figref{fig:flow} shows this process in detail, and how it can be used to generate a bivariate map for presenting uncertainty. JavaScript code for generating VSUPs given arbitrary D3~\cite{bostock2011d3} quantitative scale functions is available at \url{URL-REMOVED-FOR-REVIEW}.

Correll \ea~\cite{correll2015layercake,correll2011visualizing} use a precursor of VSUPs in their LayerCake genomics visualization tool, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``confidence fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both cases, the property of aliasing is \emph{ad hoc}, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUPs makes the choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

@@

There are multiple choices that designers must make before creating a VSUP. In particular, they must choose 1) which visual channels map to value and uncertainty, 2) a threshold of perceptual distinguishability (and, implicitly, a corresponding metric), and 3) a degradation function to determine how quickly uncertain values are aliased.

@@

How quickly the number of bins decreases is another parameter of VSUPs. If this degradation is too fast, then we may only have a few discriminable levels of uncertainty. If this degradation is too slow, then we can be overwhelmed with colors. For instance, if we only decrement by one bin per level of uncertainty, there would be $1,485$ different colors in a VSUP with a tolerance threshold of $1$ unit in CIELAB space, far too much complexity for a discrete color mapping. In the examples in this paper, we halve the number of bins at each increasing level of uncertainty. When the number of initial bins at the highly certain region is a power of $2$ (as in Fig. \ref{fig:example}), this imparts a tree structure to the resulting VSUP. As uncertainty increases, bins in the previous level map to one and only one ``parent'' in the next uncertainty level, resulting in fewer crossings of color boundaries among rows.

@@

We present a set of examples showing how VSUPs can be applied to realworld datasets. The decision of VSUPs to allocate color bins asymmetrically amongst value and uncertainty regions affords greater discriminability in regions of the chart that warrant closer scrutiny, while discouraging exploration of regions with noisy or unreliable signals.

@@

Using a public data set from the U.S. Bureau of Transportation Statistics~\cite{bts}, we created a heatmap showing average flight delay for different times of the day and days of the week for U.S. carriers in January 2017. We use standard error $\left(\sigma / \sqrt{n}\right)$ as our measure of uncertainty. Starting from an minimum discriminability threshold of $18$ units in CIELAB space, we created two alternate heatmaps of this delay information. \figref{fig:airline2d} shows a traditional bivariate map, whereas \figref{fig:airlineVsum} shows a VSUP generated under the same constraints.

Both maps illustrate a similar temporal trend: flight delays are shorter towards the beginning of the day, increasing on average over time. The traditional bivariate map affords only a coarse examination of this trend: two visible ``blocks'' of color, early in the day, and later in the afternoon. The VSUP, in contrast, has sufficient color resolution to show a quasi-linear trend. The additional color resolution for uncertainty information in VSUPs also promotes caution when considering outliers: the small number of flights on Saturdays, or departing late at night, make standard error for these delays relatively high, and the connected values more suspect than the rest of the chart.

@@

Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV). Next generation sequencing (NGS) allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) affords the analysis of only a few, large hot spots. Extending the ``confidence fog'' metaphor used in viral genome browsers like LayerCake~\cite{correll2015layercake}, VSUPs (\figref{fig:viralVsum}) enable biologists to explore interesting patterns while not being distracted by noise. Smaller scale ``hot spots'' that are mapped to the same magenta color in the traditional map are visible as bright orange lines. Pink hot spots towards the end of the genome (around reference nucleotide $9,500$) that could be plausibly considered potential regions of interest in the 2D map, are encoded as highly uncertain and of ambiguous value in the VSUPs. In fact, these regions have systematically low coverage; the lack of reliable data for these regions should discourage analysts from giving their apparent pattern equal consideration from the  stronger signals earlier in the genome.

@@

We performed a crowdsourced experiment on Amazon's Mechanical Turk to evaluate the effectiveness of VSUPs for integrating uncertainty and value information in visualizations. This focus on integration meant that we limited our experimental tasks to scenarios where the participants needed to consider both value and uncertainty before making a decision. We gave participants two main tasks:

@@

For each task, participants saw one of three types of heatmaps. Either two \textbf{juxtaposed} maps side-by-side, with separate encodings for value and uncertainty, a traditional \textbf{2D bivariate} map, where value and uncertainty are orthogonally encoded, or a \textbf{VSUP}. \figref{fig:conditions} shows examples of these factor levels. Other factors included the size of the map (either a 4x4 or 8x8 grid), and (for the prediction task), the question framing (attack or defense). Secondary factors were evenly blocked across stimuli, resulting in a 3x2 within subjects design for the identification task, and a 3x2x2 within subjects design for the prediction task. Participants reported fatigue effects in piloting, and so we limited the number of replicates in this study: participants saw 18 stimuli (6 for each map type) for the identification task, and 12 (4 for each map type) for the prediction task.

In all cases, we generated color maps based on the requirement that no two colors in the resulting color scheme were closer than 18 units of Euclidean distance in CIELAB space. This threshold results in a $9$-color bivariate map and $15$-color VSUP for all color scales used in this study. Given the size of the marks used in our study, this threshold distance is sufficient for a predicted discriminability rate of over 95\%~\cite{stone2014engineering}. Other choices of threshold, and differing univariate scales, would result in different maps (see Fig.~\ref{fig:performance}).

@@

For the identification task, we gave participants bivariate representations of uncertainty, and asked them to identify a location meeting a given criteria, e.g., ``Click on the location with the greatest uncertainty.'' There were three potential criteria, mapping to the three vertices of a VSUP: 1) high value, low uncertainty, 2) low value, low uncertainty, and 3) high uncertainty. Each map contained four locations, one of each combination of maximum and minimum value and uncertainty. The correct answer was always one of these locations. Other locations in the grid were created pseudo-randomly, with less extreme ranges of uncertainty and probability. We measured performance both in terms of accuracy (did the participant select the correct point) as well as response time.

@@

%Show:
% 1) Juxtaposition makes people ignore/underweight uncertainty (even when they don't screw up the search task)
% 2) VSUPs make people, surprise surprise, bad at distinguishing highly uncertain values
% 3) But, they make uncertainty better-integrated
%    i) lower weights on uncertain values in decision-making
%    ii) a priori better discrimination

@@

Across both the identification and prediction tasks, we had several hypotheses, stemming from our belief that VSUPs promote better \emph{integration} between uncertainty and value information, and also encourage \emph{caution} by highlighting the ambiguity or untrustworthiness introduced by uncertain data. In particular:
\begin{enumerate}
	\item Participants would be \textbf{faster and more accurate} when completing the identification task using a VSUP as compared to the other map types. Juxtaposed maps, by adding an additional search task to the lookup task, would have especially poor performance compared to VSUPs.
	\item Participants would choose targets with \textbf{less uncertainty} in the prediction task when using a VSUP compared to the other maps types. This would result in a tradeoff where they would also choose targets with \textbf{less expected value} than the other map types. Again, juxtaposed maps, by increasing the difficulty of retrieving both variables simultaneously, would highlight this difference.
\end{enumerate}

@@

For the identification task, our primary measures of performance were accuracy and response time. Our hypothesis was that VSUPs would exhibit better performance than the other charts.

We performed a one-way ANOVA investigating the effect of chart type (juxtaposed, full bivariate, or VSUP) on accuracy. We did not find a significant effect ($F(2,859) = 2.4$, $p=0.09$), although VSUPs had the highest accuracy. \figref{fig:accuracy1} illustrates this result.

@@

For the prediction task, where there is no dominating strategy, our primary measure was the distribution of guesses, in terms of both their average value, and their average uncertainty. A risky player would choose guesses with high value, regardless of the uncertainty of those points. A more conservative guesser might eschew high-risk, high-reward locations, resulting in a lower average value of guesses, but also lower uncertainty. Our hypothesis is that VSUPs, by communicating uncertainty information at the expense of value information, would promote more conservatives guesses than the other chart types.
